Implicit yaddayadda

### 2011-10-25 papers/PLDI2012/implicitization.tex
uncomparable is correct

This formula contains Boolean \emph{reachability predicates} $b_i$ for each
control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
that a path 
$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
between two points $p_{i_1}, p_{i_n} \in P_R$ 
can easily be expressed as the
cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
a \emph{destination} point, with only incoming transitions, so that the
resulting graph is acyclic and there is no paths going through control
points in $P_R$.

@@

Since we do not work on the entire transition graph but on a subset of it, we
cunjunct the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

We distinguish two phases when analysing a subgraph $P$:
\begin{enumerate}
\item We search for paths starting at point $p_i \in P_R$, such that these paths
	are in $P$, i.e are included in the working subgraph. Each time we find a
	path, we update the abstract value of the destination point of the path.
	This is the phase explained in \ref{subsec:ascending}.
\item  When the previous phase terminates, there is no more paths in
$P$, starting at $p_i$, 
that make the abstract values of the successors grow (otherwise, the SMT
solver would not have answered ``\emph{unsat}''). Narrowing iterations preserves
this property. 
However, there may exist such paths in the entire multigraph, that are not in
$P$. The second phase
computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in
algorithm~\ref{algo:combined}). We also use SMT solving to discover
these new paths, but we slightly change the SMT formula given to the SMT solver:
we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
the paths given by the SMT solver won't be in $P$.
\end{enumerate}

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionnally, the path that goes from $p_5$ to
		$p_11$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7,8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_11$ with an
		image outside of $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_11$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

@@

\begin{table*}
\begin{center}
\setlength{\tabcolsep}{1ex}
\begin{tabular}{|l
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{G/S}}
& \multicolumn{3}{c|}{\textbf{PF/S}}
& \multicolumn{3}{c|}{\textbf{PF/G}}
& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
& \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.} \\
 \hline
 \input{table1}
\end{tabular}
\end{center}
\caption{Result of the comparison of the various techniques described in this
paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
\emph{Path-focusing} (PF), and our combined technique (G+PF). For
instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
over the classic Abstract interpretation algorithm.
The $\subseteq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
$\supseteq$ the percentage of invariants better with the right-side technique,
and ``inc.'' gives the percentage of invariants that are incomparable, i.e
neither greater nor smaller.}
% TODO: le reste c'est quand c'est egal?
\label{tab:techniques}
\end{table*}

@@

While experiencing with techniques that use SMT-solving, we encountered some
limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
the SMT-solver is not able to decide the satisfiability of some SMT-formulae
expressing the semantics of non-linear programs. 
In this case, we skipped the function for which the SMT-solver returned the
``unknown'' result.
This limitation occured very rarely in our experiments, except for the analysis
of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
In such cases, such a technique should make the SMT-formula simpler to solve by using
for instance \emph{linearization}. However, such transformations are not
implemented in our analyzer.

>>>>

This formula contains Boolean \emph{reachability predicates} $b_i$ for each
control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
that a path 
$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
between two points $p_{i_1}, p_{i_n} \in P_R$ 
can easily be expressed as the
conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
a \emph{destination} point, with only incoming transitions, so that the
resulting graph is acyclic and there is no paths going through control
points in $P_R$.

@@

Since we do not work on the entire transition graph but on a subset of it, we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

We distinguish two phases when analyzing a subgraph $P$:
\begin{enumerate}
\item We search for paths starting at point $p_i \in P_R$, such that these paths
	are in $P$, i.e are included in the working subgraph. Each time we find a
	path, we update the abstract value of the destination point of the path.
	This is the phase explained in \ref{subsec:ascending}.
\item  When the previous phase terminates, there is no more paths in
$P$, starting at $p_i$, 
that make the abstract values of the successors grow (otherwise, the SMT
solver would not have answered ``\emph{unsat}''). Narrowing iterations preserves
this property. 
However, there may exist such paths in the entire multigraph, that are not in
$P$. The second phase
computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in
algorithm~\ref{algo:combined}). We also use SMT solving to discover
these new paths, but we slightly change the SMT formula given to the SMT solver:
we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
the paths given by the SMT solver won't be in $P$.
\end{enumerate}

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_11$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7,8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_11$ with an
		image outside of $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_11$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

@@

\begin{table*}
\begin{center}
\setlength{\tabcolsep}{1ex}
\begin{tabular}{|l
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{G/S}}
& \multicolumn{3}{c|}{\textbf{PF/S}}
& \multicolumn{3}{c|}{\textbf{PF/G}}
& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
& \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table1}
\end{tabular}
\end{center}
\caption{Result of the comparison of the various techniques described in this
paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
\emph{Path-focusing} (PF), and our combined technique (G+PF). For
instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
over the classic Abstract interpretation algorithm.
The $\subsetneq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
$\supsetneq$ the percentage of invariants better with the right-side technique,
and ``unc.'' gives the percentage of invariants that are uncomparable, i.e
neither greater nor smaller.}
% TODO: le reste c'est quand c'est egal?
\label{tab:techniques}
\end{table*}

@@

While experiencing with techniques that use SMT-solving, we encountered some
limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
the SMT-solver is not able to decide the satisfiability of some SMT-formulae
expressing the semantics of non-linear programs. 
In this case, we skipped the function for which the SMT-solver returned the
``unknown'' result.
This limitation occurred very rarely in our experiments, except for the analysis
of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
In such cases, such a technique should make the SMT-formula simpler to solve by using
for instance \emph{linearization}. However, such transformations are not
implemented in our analyzer.
